Quantifying unmet ICU beds in tertiary hospitals in urban India using Propensity

### 2022-09-27 manuscript.Rmd
originaldatan untampered

---
title: 'Quantifying unmet ICU beds in tertiary hospitals in urban India using Propensity
  Score Matching '
author: "Maria Eriksson"
output:
  html_document: default
  word_document: default
bibliography: bibliography.bib
csl: vancouver.csl
---

@@

Abbreviations
============
ACS - American college of surgeons
ER - Emergency room
HIC - High-income countries
LMIC - Low- and middle-income countries
PSM - propensity score matching
RCT - Randomized case-control study
TITCO -Towards Improved Trauma Care Outcomes 
TQIP - Trauma Care Improvement Program

@@

Trauma globally and in India
----------------------------
Trauma is a substantial global health problem that disproportionately affects low and middle-income countries (LMIC), resulting in significant mortality and morbidity @traumaLMIC. Despite having only 1% of the global vehicle fleet, a disproportionate 10% of global traffic casualties occur in India @indiantrauma. The annual mortality from injury in India amount to one million casualties @onemilliondead. Half of trauma mortality in India occur in-hospital, and in-hospital trauma mortality in Indian hospitals was double that of high-income countries @indiantrauma.

Mortality due to accidents in India amounted to 421 104 people in 2019, which is 31.5 per 100.000 inhabitants. The traumatic mechanisms of injury from accidents covered in this thesis are road accidents, falls, assault and burns and these make up about half of the accident mortality @nationalcrime.

@@

Health care professionals in many LMIC testify that their medical assessment oftentimes is that a trauma patient should be admitted to the ICU, but are not because such resources are scarce, overloaded and unavailable. Their medical judgement is that the lives of these patients were salvageable, and that these patients died as a direct result of not being admitted to the ICU @stuffLMICdoctorssay. Globally, an estimated two million lives could be saved annually, if the appropriate and timely trauma care provided by hospitals in many high-income countries were available and provided in low to-middle-income countries (LMICs) @livessaved

A diverse mix of public, for-profit and not-for-profit private actors provide health services in India, where private providers dominate. According to a WHO 2022 report, India has 76 709 non-ventilator supported and 39 476 ventilator-supported ICU beds @ICUbeds.

Methods to examine trauma care and ensure equivalence
-----------------------------------------------------
In order to improve and ensure equivalence of intra-hospital care, trauma quality improvement programs (TQIP) have been developed. The pilot and most notable example is the TQIP developed by the American College of Surgeons ACS. Since the implementation of the ACS pilot, TQIP has evolved to a broad term that include many kinds of processes to improve the quality of trauma care, such as mortality and morbidity conferences, using audit filters or comparing risk adjusted outcomes over time and between centres. The aim is to find discrepancies in treatment for patients presenting with similar conditions. TQIPs aim to ensure patients get equivalent care within and across hospitals, to benchmark performance of health care providers versus their peers and to improve trauma care by developing best practice guidelines @tqip.

@@

@@

<!-- Cross-sectional study: Give the eligibility criteria, and the
sources and methods of selection of participants -->
Ethical considerations
----------------------

This study does not requlre ethical permission, it refers to data collected and shared openly.

@@

Anonymization
-------------
Hospital identification number
Hospital name has been replaced with a identification number, and the conversion key has not been saved so there is not possible to reverse the conversion. Despite this, a high degree of anonymization can not be expected because the hospitals do have some distinguishable features and these are reflected in the data collected. For example,  70 % of the railroad accidents in the data set is in one of the hospitals, and one of the hosapitals is in close proximity to a big railroad hub.

Patient identification numbers
Patient names were never recorded, patients were identified by a identification number in the original unencrypted data set. In the anonymization process the identification number has been replaced with another random identification number, and the conversion key has not been saved so there is not possible to reverse the conversion.

Date observations
Dates recorded in the TITCO-I data set were shifted using a random offset of somewhere between 0 and 5 weeks for each patient, but without losing any relevant information. The offset was designed to not dissipate details that could be of potential use for researchers such as day of week and rough indication of season. The internal relations between dates have been preserved so that the timing of all recordings and discharge in chronological relation to admission are all intact.

@@

@@

Age
Very old patients are few, they have been grouped together to prevent identification via the age parameter.

@@

datatvätt:
FRÅGA: quantitativa variablerna har många NA som jag tror egentligen är NO, ex om pat har haft syrgas vid spO2 mätning. kan jag tänka lite fritt här när det känns uppenbart eller ska jag förankra? också finns det ex obs där numeriska värdet för spO2 saknas men ändå data yes/no om pat fått syrgas och jag vet inte om sån data är meningsfull?

@@

<!-- Describe any efforts to address potential sources of bias -->
The TITCO dataset are observational data collected from a clinical environment and not an experimental setting.  Medical interventions, examinations and ICU admission are not randomized, they are a consequence of a patients medical condition. There exists a causal effect between the independent and the dependent variables: a patient is sent to the ICU precisely because of the severity of their condition.

There exist a potential bias in the routines, resources and patient makeup of the participating hospitals. Becasue they are all tertiary hospitals, about two thirds of the patients have been referred from other health care providers, meaning that there is a selection bias in the patient populations. Further, hospitals with dedicated trauma wards have a propensity to receive transferred patients or police escorted patients with worse condition and prognosis.

As stated in chapter Settings, most patients have low socioeconomic status. Low socioeconomic status is associated with higher level of malnutrition and comorbidities and would have a negtive effect on vitals and the capacity of patients to tolerate certain interventions. This circumstance doesn't affect the analysis and internal validity, but reduce the generalizability of the results. 
<!-- (ÅTERKOM: har inte råd att refereas?) -->

Because the purpose of this thesis is to statistically assess how many patients that should have gone to the ICU but did not, these particular selection biases are not expected to skew the outcome, and therefore does not have to be adressed.

@@

Propensity score matching in itself introduces a source of bias from unfit assumptions. The terminology and potential sources of bias are explained chapter Study Design. In choosing predictor variables the researcher could introduce an overdetermined set and a self-fulfilling prophecy. For this reason the dataset shall be randomly split into two parts of 40% and 60% of the patients. The 60% shall be used to produce the propensity score matching, and the remaining 40% shall be used to verify it.

@@

<!-- Explain how the study size was arrived at -->
The data of 16000 trauma patients were collected at the participating hospitals over 2.5 years, as described in chapter Settings.

@@

The theory being examined in this masters thesis is that there exist a subset of the TITCO dataset that did not go to the ICU, but are very similar to the patients that did go to the ICU, in the covariates that caused ICU-treated patients to receive this intervention. The thesis is that these patients were not admitted to the ICU only for lack of such resources, not by a medical assessment.

Translated into this context this means that in this data set exist a selection bias, there is causality between the independent variables medical conditions and interventions and the dependent variable ICU admittance. Patients were admitted to the ICU as a consequence of their medical condition. Performing a randomized case-control study and randomly assigning or withholding a live-saving treatment would obviously be immoral, but we believe that in the TITCO data set exist patients who were not admitted into the ICU by volition but for scarcity of such resources. Such a group would be akin to an non-experimental control group and PSM can be used to identify and assemble it.  If such a subset can be identified by PSM, then it naturally follows that there exists shortage of ICU beds in tertiary hospitals in India. It also follows that the shortage can be quantified by the proportionality of this subset versus the subset that did receive ICU care. If, for every patient that did receive ICU care, there was another patient with similar predictors that wasn’t admitted due to scarcity of ICU resources, that would mean the ICU resources are 50% of what they ought to be.

The propensity score model is developed using the Training Data and tested using the Validation Data. The workflow contain a series of steps and the work flow should understood as linear, but iterative. The output of each step is the input for later steps. Therefore the output of the last step reflect the cumulative error from unfit assumptions made the second and fourth step. The output is tested against Validation Data. By reverting back and adjusting assumptions in an iterative process, the model is continuously improved and the accuracy and validity is improved.

The first step of PSM is to randomly split the data set into two parts, 60% and 40%. The 60% part is a Training Part used to delvelop and train the PSM, and the 40% part is the Validation Part. This is done because the inherent bias of the PSM method described in chapter Bias. Training Data is used in steps 2-4 and the output is then validated using Validation Data.

Second step is to identify predictor variables: covariates most predictive of the dependent variable, as described in chapter Predictor selection. In this context it means to first identify the conditions and medical interventions received that correlate best to being admitted to ICU care. This selection is based on clinical experience as well as analysis of the Training Data. In the data analysis all properties of the patients that were sent to the ICU will be compared to the patents who did not, in order to identify properties that distinguishes this cohort., Familiarity with the collection of source data is crucial to minimize the risk to choose irrelevant predictors and introduce error.

@@

Fourth step is PSM. In PSM members of the subset that did not get ICU care but had similar predictors - and therefore similar propensity scores - as the subgroup that did get ICU care, are matched to patients who did go to the ICU. There are a few different ways to perform this matching but they are all imperfect because propensity scores are unique and perfect matching is unattainable There are control mechanisms and limitations but a risk of applying too fix or too loose matching criteria the outcome could inadvertedly vastly under- or overestimate the  need of ICU beds remain.

The same procedure is repeated on the Validation Data using the same predictors and the same matching criteria, and this should produce a cohort comparable to the one trom Training Data.

(b) Describe any methods used to examine subgroups and interactions

@@

Predictor variables will be selected using both clinical judgement and data analysis.

Clinical experience compels inclusion of vital parameters, irregardless if there is any statistical motivation. Such variables are heart rate, systolic blood pressure and respiration rate.

The TITCO-I data set is analyzed mathemathically in pursuit of suitable predictor variables. The data set is divided into two subgroups, the cohort that were admitted to the ICU and the cohort that was not. In comparison between these subgroups, identify variables that distinguishes the former from the latter. The quantitative and qualitative properties were considered separately.

Mean and standard deviation for the quantitative data is analysed for the ICU and non-ICU cohort respectively. p- and t-values when comparing these cohorts to show if there are any statistically significant differences.

For the qualitative data, the propensity of each parameter to get a patient admitted to the ICU is in relation to the proportion of patients that were admitted to the ICU from the whole TITCO-I data set, about a third. If considerably more than a third of each qualitative parameter are admitted to the ICU, then that parameter is of interest.

Some qualitative parameters such as intubation are recorded after both 1 and 24 hours, if applicable. The record at 1 hour has two options indicating the patient is intubated, "before arrival" meaning the patient was intubated at a their first health care facility and "yes" meaning the patient was intubated within the first hour. Because these two options both mean the patient was intubated after 1 hour, an aggregate parameter has been created for this, and a further parameter "intubated after 24 hours" that summarize the patients who were intubated on arrival, the patients who were intubated within the first hour and the patients who were intubated between 1 and 24 hours.

FRÅGA: det finns pat som är intuberade "before arrival" men som inte är noterade som transferred från ett annat sjukhus. De har kommit med motor rickshaw. Är det rimligt att anta att de är transferred från ett sjukhus iallafall för vem har annars intuberat dessa patienter?

<!-- 
The parameter "taken to the operation theatre" is a heterogenous qualitative parameter, where only a small proportion of patients had been taken within the first hour but a substantial proportion of patients had been taken to the operation theatre after 24 hours. There is a need to consider the free text field operation type to filter out for operation types with higher propensity for ICU admission from operations with lower propensity. 
-->
The parameter "taken to the operation theatre" is a heterogenous qualitative parameter, and operations performed in the first hour tend to be more acute than operations performed within the first day of hospitalization. There is a need to consider the free text field operation type to filter out for operation types with higher propensity for ICU admission from operations with lower propensity. 
FRÅGA: borde man ta operation within 24 hours och komplettera med text based operationstyp?

@@

@@

```{r}
setwd("C:/Users/maria/Downloads/Packt Learning RStudio for R Statistical Computing 2012 RETAIL eBook-repackb00k/unmet-ICU-beds")
load("table3.rdata")

knitr::kable(table3, caption = "Means and standard devations of quantitative variables of ICU and non-ICU admitted patients, respectively")
```

@@

P-values and T-values of non-ICU vs ICU patients:
```{r}
setwd("C:/Users/maria/Downloads/Packt Learning RStudio for R Statistical Computing 2012 RETAIL eBook-repackb00k/unmet-ICU-beds")
load("statresult.rdata")

knitr::kable(statresult, caption = "P- and t-values of the quantitative values in comparison of non-ICU and ICU patients")
```
FRÅGA: är p-värdena rimliga tycker du? de är superlåga och min känsla - med 5 hp statistik i ryggen - är att std-avvikelsen är för stor för det?

>>>>

---
title: 'Quantifying unmet ICU beds in tertiary hospitals in urban India using Propensity
  Score Matching '
author: "Maria Eriksson"
output:
  pdf_document: default
  html_document: default
  word_document: default
bibliography: bibliography.bib
csl: vancouver.csl
---

@@

Abbreviations
============
ACS - American college of surgeons

ER - Emergency room

HIC - High-income countries

LMIC - Low- and middle-income countries

PSM - propensity score matching

RCT - Randomized case-control study

TITCO -Towards Improved Trauma Care Outcomes

TQIP - Trauma Care Improvement Program

@@

Trauma globally and in India
----------------------------
Trauma is a substantial global health problem that disproportionately affects low and middle-income countries (LMIC), resulting in significant mortality and morbidity @traumaLMIC. 4.8 million people die from trauma annually, 90% are in LMIC and 20% in India alone @onemilliondead. Despite having only 1% of the global vehicle fleet, a disproportionate 10% of global traffic casualties occur in India @indiantrauma. Half of trauma mortality in India occur in-hospital. There is no national trauma registry in India, but isolated studies have shown that in-hospital mortality within 30 days of hospitalization is higher than similarly injured patients in HIC hospitals @indiantrauma. Globally, an estimated two million lives could be saved annually, if the appropriate and timely trauma care provided by hospitals in many high-income countries were available and provided in low to-middle-income countries (LMICs) @indiantrauma.

Health care professionals in many LMIC testify that their medical assessment oftentimes is that a trauma patient should be admitted to the ICU, but are not because such resources are scarce, overloaded and unavailable. Their medical judgement is that the lives of these patients were salvageable, and that these patients died as a direct result of not being admitted to the ICU @stuffLMICdoctorssay.

@@

A diverse mix of public, for-profit and not-for-profit private actors provide health services in India, where private providers dominate. According to a WHO 2022 report, India has 76 709 non-ventilator supported and 39 476 ventilator-supported ICU beds @ICUbeds. Because of the type and severity of injuries, ICU resources are often essential in care of trauma patients @titco.

<!-- KommentarJohanna: I suggest adding more details about the context in India, the health care system, how common trauma is in India, how many ICU beds there are, why ICU care is important for the trauma population and so on.. 
-->
<!-- KommentarMartin:You can definitely write more about the health care system in India, the trauma system, the availability of ICUs, how common ICU care is in trauma, why trauma patients need ICU care, how long they stay, outcomes etc -->

<!-- Respons: Ja det är magert men det finns en orsak till det, jag har undvikit detta av två skäl
 1.Datan som står att finna om ICU vård resurserna i Indien är lite spretig, men enligt WHO så är den runt 8 per 100.000 och andra trovärdiga bedömare är där nånstans +- 50%. Det är i paritet med Sverige? Förklaringen står kanske att finna i att definitionen av en ICU bed inte är vedertagen men trots mycket läsande innantill har jag inte kunnat avfärda eller förklara detta. 
2. pkt 1 lyfter en fråga som ökar komplexiteten: det finns lite olika levels av ICU beds, vilken är det vi menar? Eftersom statistiken på ICU resurser är dåligt differentierad och TITCO datasetet inte specificerar det, ska frågan undvikas för att inte öka komplexiteten för sig själv? -->

Methods to examine trauma care and ensure equivalence
-----------------------------------------------------
In order to improve and ensure equivalence of intra-hospital care, trauma quality improvement programs (TQIP) have been developed. The pilot and most notable example is the TQIP developed by the American College of Surgeons ACS. Since the implementation of the ACS pilot, TQIP has evolved to a broad term that include many kinds of processes to improve the quality of trauma care, such as mortality and morbidity conferences, using audit filters or comparing risk adjusted outcomes over time and between centres. The aim is to find discrepancies in treatment for patients presenting with similar conditions. This difference in in-hospital mortality is likely multi-factorial, and any factors have been proposed to explain it such as differences in triage and delay-of-care, as well as shortage of hospital resources including ICU beds @onemilliondead ,@traumaLMIC . 
TQIPs aim to ensure patients get equivalent care within and across hospitals, to benchmark performance of health care providers versus their peers and to improve trauma care by developing best practice guidelines @tqip.

@@

<!-- Ni vil båda flytta fokus från detta, och jag förstår inte riktigt varför det är med kopplingen är luddig. Är det för att säga: det finns TQIPs som ger en fingervisning om hur ICU ska dimensioneras, bestyckas, bemannas och bedrivas Men det finns inga färdiga TQIPs som kan användas för att kvantifiera ICU-bristen i Indien och vi tror det är vanskligt att använda de befintliga eftersom de skapats under så olika förhållanden. Och därför måste vi börja från noll? -->

@@

<!-- Cross-sectional study: Give the eligibility criteria, and the
sources and methods of selection of participants -->

Ethical considerations
----------------------

This study does not require ethical permission, it refers to data collected and shared openly.

@@

Anonymization
-------------
Hospital identification number

Hospital name has been replaced with a identification number, and the conversion key has not been saved so there is not possible to reverse the conversion. Despite this, a high degree of anonymization can not be expected because the hospitals do have some distinguishable features and these are reflected in the data collected. For example,  70 % of the railroad accidents in the data set is in one of the hospitals, and one of the hosapitals is in close proximity to a big railroad hub.

Patient identification numbers

Patient names were never recorded, patients were identified by a identification number in the original unencrypted data set. In the anonymization process the identification number has been replaced with another random identification number, and the conversion key has not been saved so there is not possible to reverse the conversion.

Date observations

Dates recorded in the TITCO-I data set were shifted using a random offset of somewhere between 0 and 5 weeks for each patient, but without losing any relevant information. The offset was designed to not dissipate details that could be of potential use for researchers such as day of week and rough indication of season. The internal relations between dates have been preserved so that the timing of all recordings and discharge in chronological relation to admission are all intact.

@@

Age

Very old patients are few, they have been grouped together to prevent identification via the age parameter.

@@

@@

@@

<!-- Describe any efforts to address potential sources of bias -->
The TITCO dataset are observational data collected from a clinical environment and not an experimental setting.  Medical interventions, examinations and ICU admission are not randomized, they are a consequence of a patient's medical condition. There exists a causal effect between the independent and the dependent variables: a patient is sent to the ICU precisely because of the severity of their condition.

There exists a potential bias in the routines, resources and patient makeup of the participating hospitals. Because they are all tertiary hospitals, about two thirds of the patients have been referred from other health care providers, meaning that there is a selection bias in the patient populations. Further, hospitals with dedicated trauma wards have a propensity to receive transferred or police escorted patients with worse condition and prognosis.

As stated in chapter Settings, most patients have low socioeconomic status. Low socioeconomic status is associated with higher level of malnutrition and comorbidities and would have a negative effect on vitals and the capacity of patients to tolerate certain interventions. This circumstance doesn't affect the analysis and internal validity, but reduce the generalizability of the results. 
<!-- (ÅTERKOM: har inte råd att refereas?) -->

Because the purpose of this thesis is to statistically assess how many patients that should have gone to the ICU but did not, these particular selection biases are not expected to skew the outcome, and therefore does not have to be addressed.

@@

<!--  
Propensity score matching in itself introduces a source of bias from unfit assumptions. The terminology and potential sources of bias are explained chapter Study Design. In choosing predictor variables the researcher could introduce an overdetermined set and a self-fulfilling prophecy. For this reason the dataset shall be randomly split into two parts of 40% and 60% of the patients. The 60% shall be used to produce the propensity score matching, and the remaining 40% shall be used to verify it. 
-->

<!-- Fråga: vad är distinktionen mellan bias och error med PSM egentligen? är det error eller bias som uppstår om jag ex skulle välja ICU-sjukhuset som prediktor? Är det bias att ha för generösa calipers? Som jag förstår finns risk att få vilket resultat som helst med val metod för matching och limitations ? Och att jag väljer calipers som producerawr resultat som jag tycker känns "rimliga"? Ex jag kanske väntar mig att ICU bristen är 20% för jag föreställer mig att de patienter jag söker är de som är "på ribban" alltså de matchar de "friskaste" 20% av dem som var på ICU. Det känns mindre sannolikt att en patient inte fått ICU om den är jämförbar med någon som var en av de mest skadade 20% på ICU.   -->

@@

<!-- Explain how the study size was arrived at -->
Data of 16000 trauma patients were collected at the participating hospitals over 2.5 years, as described in chapter Settings.

@@

The theory being examined in this master's thesis is that there exist a subset of the TITCO dataset that did not go to the ICU, but are very similar to the patients that did go to the ICU, in the covariates that caused ICU-treated patients to receive this intervention. The thesis is that these patients were not admitted to the ICU only for lack of such resources, not by a medical assessment.

Translated into this context this means that in this data set exist a selection bias, there is causality between the independent variables medical conditions and interventions and the dependent variable ICU admittance. Patients were admitted to the ICU as a consequence of their medical condition. Performing a randomized case-control study and randomly assigning or withholding a live-saving treatment would obviously be immoral, but we believe that in the TITCO data set exist patients who were not admitted into the ICU not by volition but for scarcity of such resources. Such a group would be akin to an non-experimental control group and PSM can be used to identify and assemble it.  If such a subset can be identified by PSM, then it naturally follows that there exists shortage of ICU beds in tertiary hospitals in India. It also follows that the shortage can be quantified by the proportionality of this subset versus the subset that did receive ICU care. If, for every patient that did receive ICU care, there was another patient with similar predictors that wasn’t admitted due to scarcity of ICU resources, that would mean the ICU resources are 50% of what they ought to be.

The propensity score model is developed using the Training Data and tested using the Validation Data. The workflow contains a series of steps and the work flow should understood as linear, but iterative. The output of each step is the input for later steps. Therefore the output of the last step contain the cumulative error from unfit assumptions made the second and fourth step. The output is tested against Validation Data. By reverting back and adjusting assumptions in an iterative process, the model is continuously improved and the accuracy and validity is improved.

<!-- ÅTERKOM inte klart bara gissat -->

The first step of PSM is to randomly split the data set into two parts, 60% and 40%. The 60% part is a Training Part used to develop and train the PSM, and the 40% part is the Validation Part. This is done because the inherent bias of the PSM method described in chapter Bias. Training Data is used in steps 2-4 and the output is then validated using Validation Data.

Second step is to identify predictor variables: covariates most predictive of the dependent variable, as described in chapter Predictor selection. In this context it means to first identify the conditions and medical interventions received that correlate best to being admitted to ICU care. This selection is based on clinical experience as well as analysis of the Training Data. In the data analysis all properties of the patients that were sent to the ICU will be compared to the patents who did not, in order to identify properties that distinguishes this cohort. Familiarity with the collection of source data is crucial to minimize the risk to choose irrelevant predictors and introduce error.

@@

Fourth step is PSM. In PSM members of the subset that did not get ICU care but had similar predictors - and therefore similar propensity scores - as the subgroup that did get ICU care, are matched to patients who did go to the ICU. There are a few different ways to perform this matching but they are all imperfect because propensity scores are unique and perfect matching is unattainable There are control mechanisms and limitations but a risk of applying too fix or too loose matching criteria the outcome could inadvertently  vastly under- or overestimate the  need of ICU beds remain.  
The same procedure is repeated on the Validation Data using the same predictors and the same matching criteria, and this should produce a cohort comparable to the one from Training Data. The proportion of this subset versus the subset that did receive ICU care is also the proportion of ICU bed shortage. If, for every patient that did receive ICU care, there was another patient with similar predictors that wasn’t admitted due to scarcity of ICU resources, that would mean the ICU resources are 50% of what they ought to be.

<!-- (b) Describe any methods used to examine subgroups and interactions -->

@@

Predictors are independent variables with strong causality with the dependent varaiable. Predictor variables will be selected using both clinical judgement and data analysis. Because causality is presumed, predictors must precede ICU admission. For example, though there is a strong correlation between ICU admission and mortality, mortality can not be a predictor because it did not cause ICU admission.

Clinical experience compels inclusion of vital parameters, regardless if there is any statistical motivation. Such variables are heart rate, systolic blood pressure and respiration rate.

The TITCO-I data set is analysed mathematically  in pursuit of suitable predictor variables. The data set is divided into two subgroups, the cohort that were admitted to the ICU and the cohort that was not. In comparison between these subgroups, identify variables that distinguishes the former from the latter. The quantitative and qualitative properties were considered separately.

Mean and standard deviation for the quantitative data is analysed for the ICU and non-ICU cohort respectively. p- and t-values when comparing these cohorts to show if there are any statistically significant differences. It is plausible to construct new variables from quantitative data, such as trends.

Timer data can be used to construct quantitative variables delay-of-care. Second and third delays, corresponding to the delays between incident and hospital admission and between admission and first survey have been proposed to have high relevance for patient outcome @indiantrauma .

For the qualitative data, the propensity of each parameter for ICU admission is compared to the proportion of patients that were admitted to the ICU from the whole TITCO-I data set, about a third. If considerably more than a third of each qualitative parameter are admitted to the ICU, then that parameter is of interest. It has been relevant to combine qualitative parameters into aggregate qualitative varaibles. For example, the variable intubation at 1 hour has two options indicating the patient is intubated: "before arrival" meaning the patient was intubated at a their first health care facility and "yes" meaning the patient was intubated within the first hour. Because these two options both mean the patient was intubated after 1 hour, an aggregate parameter has been created for this. Because the data set also contain information of whether a patient was intubated within the first and 24th hour the parameter "intubated after 24 hours" were created that summarize the patients who were intubated on arrival, the patients who were intubated within the first hour and the patients who were intubated between 1 and 24 hours.

@@

It is also plausible to include aggregate variables, by combining unrelated variables of the same type or across varaible types. Selection and construction of such aggregate parameters to be explored will be based on clinical experience not data analysis. An example of this is the heterogenous qualitative parameter "taken to the operation theatre". Many operation types are strongly correlated to ICU admission - especially if they were performed within the first hour of admission - but most of the operations performed are lower priority orthopedic operations with no correlation to ICU admission. There is a need to combine operation timing and free text fields "operation type" and "intraoperative findings" to filter out operations with higher propensity for ICU admission.
<!-- FRÅGA prediktorval: borde man ta operation within 24 hours och komplettera med text based operationstyp?
-->

@@

```{r}
setwd("C:/Users/maria/Downloads/Packt Learning RStudio for R Statistical Computing 2012 RETAIL eBook-repackb00k/unmet-ICU-beds")
load("table5.rdata")

knitr::kable(table5, caption = "Means, standard devations of quantitative variables of ICU and non-ICU admitted patients, respectively and p- and t-values of comparison between the sets")
```
<!-- FRÅGA: är p-värdena rimliga tycker du? de är väldigt låga och min känsla - med 5 hp statistik i ryggen - är att std-avvikelsen är för stor för det? -->

@@
